\section*{Incubation}
\addcontentsline{toc}{section}{Incubation}
\setcounter{page}{1}

The Internet's massive size and popularity today make it hard to imagine its
humble beginning. What we now know as the Internet began as a research project
sponsored by the Defense Advanced Research Project Agency (DARPA) called the
Arpanet. On the day of its birth in December, 1969, the Arpanet consisted of
only four connected nodes at the University of California in Los Angeles, the
Stanford Research Institute, the University of California in Santa Barbara, and
the University of Utah.
Lines of communication between these nodes were ``two million times slower than
today's fastest networks.''\cite{strawn_masterminds_2014} 

The Arpanet opened the door for a flurry of new research projects. New ideas
grew into numerous applications such as file sharing, remote logon, and email.
Soon, more nodes and networks were added. By the mid 1980s, the growth and success
of the Arpanet prompted the National Science Foundation (NSF) to build a network
connecting research universities to some of its newly built supercomputer
centers.
This new network came to be known as the NSFnet.
It made internetworking a prolific research topic,
eventually connecting ``more than 2,000 universities and colleges and a
number of high tech companies.''\cite{strawn_masterminds_2014} Ultimately, the
1990s would see the NSFnet becoming commercialized into the Internet that we
know today, but not without some growing pains. The Morris
Worm\footnote{also called the Internet Worm} was one of the earliest and most
vehement growing pains the developing internet\footnote {
The above summary of the
birth of the Internet used the names ``Arpanet'', ``NSFnet'', and ``Internet''
for various stages in the growth of a widely connected network of computers. For
simplicity, the rest of this paper will collectively refer to all of these
stages as the ``internet''.
} 
would face.

An important detail in the internet's conception and early life was the
attitude of its creators.
The internet was developed by researchers focused on
optimizing its efficiency. Some security
measures were built in as an afterthought, but security was not woven into
the core fabric of the internet or its applications. As the internet
grew, researchers who used and designed
it were generally assumed to have good intentions. ``The Internet was
like a small town where people thought little of leaving their doors unlocked. Internet security was
seen as a mostly theoretical problem, and software vendors treated security
flaws as a low priority.''\cite{lee_washpost_2013} Lee\cite{lee_washpost_2013}
quotes the sentiments of Dr.
Eugene Spafford \footnote{
Dr. Spafford is a computer security researcher and professor at Purdue
University. He is also executive director of Purdue's Center for Education
and Research in Information Assurance and Security and an internationally
recognized expert in computer security.}:

\begin{quote}
The majority of people had some tie to computation for their jobs. I wouldn't
say that we trusted each other, but there was more a community sense of caring
for the stability and appropriate use of the computing systems\ldots. \\
There was no such thing as a firewall back then. You didn't have people who were
vandals or anarchists or criminals as much. 
\end{quote}

A lack of
security would ultimately provide a temperate environment for the incubation of
a virulent infection: the Morris Worm.

The remainder of this
paper tunnels through the worm hole in history that the Morris Worm
has left behind. First, it covers \textit{what} happened during the worm's
outbreak and eradication. Second, it recounts the discoveries that were made
by researchers about \textit{how} the worm works. Third,
it discusses the \textit{immediate impact} that the worm had on the research
community and the general public as they convalesced after the infection.
Fourth, it traces the worm's \textit{long term influence} on the evolution of future
worms and on shifting security perspectives. Finally, the paper
provides a post mortem synthesis on the quality and depth of the worm's overall
impact on the history of computer security. 
