%%%%%%%%%%%%% OUTBREAK %%%%%%%%%%%%%%%%%%%
\section*{Outbreak}
\addcontentsline{toc}{section}{Outbreak}
This section describes events that took place between November $3^{rd}$ and
Novermber $8^{th}$ of 1988. It outlines the Morris Worm's release, spread, and
eventual eradication at a high level.
Many works covering the Morris Worm begin with some variation of the phrase
``On the evening of November $2^{nd}$, 1988\ldots''.
 \cite{seeley_tour_1989}
 \cite{spafford_crisis_1989}
 \cite{lee_washpost_2013}
 \cite{spafford_internet_1989-1}
 \cite{spafford_internet_1989}
 \cite{eichin_microscope_1989}
To honor this tradition, this section does the same: 

%%%%%%%%%%%%% BLACK THURSDAY %%%%%%%%%%%%%%%%%%%
\subsection*{Black Thursday}
\addcontentsline{toc}{subsection}{Black Thursday}

On the evening of Wednesday November $2^{nd}$, 1988, Robert Tappan Morris, a
first year graduate student at Cornell University, released a worm onto the
internet. The worm was released at around 6 pm at MIT. Spreading rapidly, by 11
pm it had infected machines at
the University of Pittsburgh,
RAND Corporation in Santa Monica,
UC Berkely,
the University of Maryland,
the University of Utah,
Stanford,
the University of Minnesota,
and the University of North Carolina
\cite{seeley_tour_1989}
\cite{spafford_internet_1989-1}.

Perhaps the worm spread faster than even Morris had anticipated. Around
11 pm and again at around 11:30 pm, he contacted Andrew Sudduth and Paul Graham
\footnote{
Andrew Sudduth was a friend of Morris'. He and Paul Graham were both
members of the technical staff at Harvard University's Aiken Computational
Laboratory \cite{lee_washpost_2013}
}
to tell them that he had released the worm and steps that could be taken to stop
it. Furthermore, Morris requested that Andrew Sudduth alert the research
community of the presence of the worm and how it might be stopped, which
Sudduth did anonymously via email to a widely used internet research mailing
list, called TCP-IP shortly thereafter.
Unfortunately, by that time system administrators had already noticed the worm and had shut off
internet gateways in an effort to quarantine the infection; thereby blocking
Sudduth's email for several days
\cite{lee_washpost_2013}\cite{eisenberg_cornell_1989}.

The worm continued to spread throughout the night. By early Thursday morning,
the infection had spread to the University of Arizona, Princeton University,
Lawrence Livermore Labs, UCLA, Purdue University, Georgia Tech, Dartmouth,
the Army Ballistics Research Lab, and the University of Chicago, amongst others.
By this point many system administrators were aware of the spreading infection, and Peter Yee of UC Berkeley and NASA
Ames Research Center had posted a message to the TCP-IP mailing, stating ``We
are under attack''\cite{seeley_tour_1989}
\cite{spafford_internet_1989-1}. Eventually, Thursday November $11^{th}$ would
come to be known as ``Black Thursday''. 

At its peak, the infection is estimated to have spread to around 6,000 machines.
That is 10\% of the approximately 60,000 computers connected to the internet at
that time\cite{eichin_microscope_1989}\cite{marsan_morris_2008}. The worm
targeted systems running 4.2 or 4.3 BSD UNIX and SunOS. The worm
caused many of its infected hosts to crash.
Hosts that didn't crash were riddled with processes that appeared to be shells.
Their process tables, open file tables, and sometims swap space were exhausted.
Latency soared as legitimate user processes competed for cpu time with worm
processes. Logs showed odd activity that wasn't actually being caused by the
users that they were recording. Finally, strange files were appearing in the
/usr/tmp directory \cite{seeley_tour_1989}\cite{spafford_crisis_1989}.

Fortunately, the research community had not only noticed the worm by early
Thursday morning, they had already begun fighting back.



%%%%%%%%%%%%% INNOCULATION %%%%%%%%%%%%%%%%%%%
\subsection*{Inoculation}
\addcontentsline{toc}{subsection}{Inoculation}

\cite{spafford_internet_1989-1} points out:
\begin{quote}
It is particularly interesting to note how quickly and how widely the Worm spread.
It is also significant to note how quickly it was identified and stopped by an ad hoc col-
lection of "Worm hunters" using the same network to communicate their results.
\end{quote}

By 5 am on Thursday, ``less than 12 hours after the program was first discovered on the network, the Computer Systems Research Group at Berkeley had
developed an interim set of steps to halt its spread''
\cite{spafford_crisis_1989}. While Black Thursday saw many a researcher and
system administrator struggle with slow and crashing machines and trickling
network connectivity, it also saw many swift and staggering successes for them.
Mailing lists, including Dr. Spaffords Phage Mailing List
\footnote{
The Phage Mailing List would become the main conduit of communication and source
of information about the worm both during the worm crisis and in subsequent weeks as Morris
was prosecuted
\cite{spafford_crisis_1989}\cite{spafford_internet_1989-1}\cite{lee_washpost_2013}.
} 
were put together and mainted to coordinate anti-worm efforts.
Researchers captured the worm and analyzed its behavior. They devoloped worm
vaccines with impressive speed. By the end of the day, software patches had been posted to
eliminate the vulnerabilities exploited by the worm, namely in the 
\textit{sendmail} and \textit{finger} applications.
\cite{spafford_crisis_1989}
\cite{seeley_tour_1989}.

``By about 9 p.m. Thursday, another simple, effective method of stopping the
invading program, without altering system utilities, was discovered at Purdue
and also widely published''\cite{spafford_internet_1989-1}. This involved simply
creating a file called \textit{sh} in the /usr/tmp directory.

Autonamously teaming up, researchers had discovered how to defeat the Morris
Worm in just one day and infection was on the decline. Over the next several
days, the Morris worm was eradicated. \cite{spafford_internet_1989-1} relates
that ``By Tuesday, November 8, most machines had connected back to the Internet
and traffic patterns had returned to near normal.''

By early Friday, November $4^{th}$, researchers had disassembled the worm's
code 
\cite{spafford_crisis_1989}
\cite{seeley_tour_1989}.
Several surprises were in store for those performing the autopsy.